\begin{hcarentry}[updated]{Haskell XML Toolbox}
\label{hxt}
\report{Uwe Schmidt}%05/10
\status{seventh major release (current release: 8.5.2)}
\makeheader

\subsubsection*{Description}

The Haskell XML Toolbox (HXT) is a collection of tools for processing XML with
Haskell. It is itself purely written in Haskell 98. The core component of the
Haskell XML Toolbox is a validating XML-Parser that supports
almost fully the Extensible Markup Language (XML) 1.0 (Second Edition).
There is a validator based on DTDs and a new more powerful one for
Relax~NG schemas.

The Haskell XML Toolbox is based on the ideas of HaXml~\cref{haxml} and HXML,
but introduces a more general approach for processing XML with Haskell.
The processing model is based on arrows. The arrow interface is more flexible
than the filter approach taken in the earlier HXT versions and in HaXml.
It is also safer; type checking of combinators becomes possible with the arrow
approach.

HXT is partitioned into 6 packages: The base package {\tt hxt}, the package for the
old approach working with filter {\tt hxt-filter} (this one will not be further
developed), the package {\tt hxt-xpath} for XPath functionality, {\tt hxt-xslt}
for the XSLT interpreter, {\tt hxt-binary} a small package for binary (de-)serialization
of HXT DOM trees, and {\tt hxt-cache}, a package for caching XML/HTML documents in parsed
format.

\subsubsection*{Features}

\begin{itemize}
\item Validating XML parser
\item Very liberal HTML parser
\item Lightweight lazy parser for XML/HTML based on Tagsoup~\cref{tagsoup}
\item Easy de-/serialization between native Haskell data and XML by pickler and pickler combinators
\item XPath support
\item Full Unicode support
\item Support for XML namespaces
\item Cabal package support for GHC
\item HTTP access via Haskell bindings to libcurl
\item Tested with W3C XML validation suite
\item Example programs
\item Relax~NG schema validator
\item An HXT Cookbook for using the toolbox and the arrow interface
\item Basic XSLT support
\item Git repository with current development versions of all packages
  \url{http://git.fh-wedel.de/repos/hxt.git}
\end{itemize}

\subsubsection*{Current Work}

Currently mainly maintenance work is done. This includes
space and runtime optimizations.

The HXT library is extensively used in the Holumbus
project~\cref{holumbus}, there it forms the basis for the index generation.
Development is currently driven by the needs of the Holumbus project.

\FurtherReading
The Haskell XML Toolbox Web page
(\url{http://www.fh-wedel.de/~si/HXmlToolbox/index.html})
includes downloads, online API documentation, a cookbook with nontrivial examples
of XML processing using arrows and RDF documents, and master theses describing the
design of the toolbox, the DTD validator, the arrow based Relax~NG
validator, and the XSLT system.

A getting started tutorial about HXT is available in the Haskell Wiki (\url{http://www.haskell.org/haskellwiki/HXT}
). The conversion between XML and native Haskell datatypes is
described in another Wiki page
(\url{http://www.haskell.org/haskellwiki/HXT/Conversion_of_Haskell_data_from/to_XML}).
\end{hcarentry}
